---
title: "Final Project"
author: "Kaustubh Patil"
date: "2024-05-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Installing packages]{.underline}

```{r}
# install.packages("readxl")
# install.packages("tseries")
# install.packages("ggplot2")
# install.packages("DataCombine")
# install.packages("xts")
# install.packages("zoo")
# install.packages("forecast")
# install.packages("stats")
# install.packages("fracdiff")
# install.packages("car")
# install.packages("MASS")
# install.packages("boot")
# install.packages("dplyr")
# install.packages("imputeTS")
# install.packages("TSA")
```

[Loading libraries]{.underline}

```{r}
library(readxl)
library(tseries)
library(ggplot2)
library(DataCombine)
library(xts)
library(zoo)
library(forecast)
library(stats)
library(fracdiff)
library(car)
library(MASS)
library(boot)
library(dplyr)
library(imputeTS)
library(TSA)
```

[Loading data]{.underline}

```{r}
vodka_shipments <- readxl::read_xlsx("C:/Users/khpat/Dropbox/ADSP 31006 Time Series Analysis & Forecasting/Final Project/Time Series Data.xlsx")
```

```{r}
vodka_consumption <- readxl::read_xlsx("C:/Users/khpat/Dropbox/ADSP 31006 Time Series Analysis & Forecasting/Final Project/TS Consumption.xlsx")
```

Extracting the date from the consumption data

```{r}
vodka_consumption$period_description_short <- sub("^.*?\\s(\\d{1,2}/\\d{2}/\\d{2})$", "\\1", vodka_consumption$period_description_short)
```

Converting the strings to date objects in shipment data

```{r}
vodka_shipments$`End of Week` <- as.Date(vodka_shipments$`End of Week`)
```

Formatting and converting the date strings to date

```{r}
vodka_consumption$period_description_short <- as.Date(vodka_consumption$period_description_short, format = "%m/%d/%y")
```

Ordering the dates

```{r}
vodka_consumption$period_description_short <- vodka_consumption$period_description_short[order(vodka_consumption$period_description_short)]
```

Calculating the liter quantities

```{r}
vodka_shipments$shipping_liters <- 9 * vodka_shipments$`9L Total`
vodka_consumption$consumption_liters <- 9 * vodka_consumption$`9L Data`
```

[Creating and plotting the datasets as time series objects]{.underline}

```{r}
vodka_shipments_xts <- xts(x = vodka_shipments$shipping_liters, order.by = vodka_shipments$`End of Week`)
plot(vodka_shipments_xts)
```

```{r}
vodka_consumption_xts <- xts(x = vodka_consumption$consumption_liters, order.by = vodka_consumption$period_description_short)
plot(vodka_consumption_xts)
```

[Plotting both time series]{.underline}

```{r}
# Merging the time series objects
combined_xts <- merge(vodka_shipments_xts, vodka_consumption_xts, join = "inner")

# Converting to data frame for ggplot
combined_df <- fortify.zoo(combined_xts)
names(combined_df) <- c("Date", "Shipments", "Consumption")

# Plot using ggplot
ggplot(data = combined_df, aes(x = Date)) +
  geom_line(aes(y = Shipments, colour = "Shipments")) +
  geom_line(aes(y = Consumption, colour = "Consumption")) +
  labs(
    title = "Vodka Shipments & Consumption Over Time",
    x = "Date",
    y = "Volume in Liters",
    colour = "Legend"
  ) +
  theme_minimal()
```

Looking at the time series plots of both the consumption and shipping data, both series appear stationary at a glance.

```{r}
adf.test(vodka_shipments_xts)
kpss.test(vodka_shipments_xts)
```

The significant Dickey Fuller test means we can reject the null hypothesis which states that the series is explosive for the shipping data. Similarly, the insignificant KPSS test p value means we fail to reject the null which states the series is stationary.

```{r}
adf.test(vodka_consumption_xts)
kpss.test(vodka_consumption_xts)
```

The significant Dickey Fuller test means we can reject the null hypothesis which states that the series is explosive for the consumption data. Similarly, the insignificant KPSS test p value means we fail to reject the null which states the series is stationary.

```{r}
acf(vodka_shipments_xts)
```

```{r}
acf(vodka_consumption_xts)
```

Very minimal significant autocorrelation of lags for both series. This indicates that an ARFIMA model would not be well suited to this data.

Merging both datasets -

```{r}
combined_data <- merge(vodka_shipments, vodka_consumption, by.x = "End of Week", by.y = "period_description_short")
```

Splitting into training and testing data

```{r}
train <- combined_data[combined_data$`End of Week` <= as.Date("2023-12-31"), ]
test <- combined_data[combined_data$`End of Week` >= as.Date("2024-01-01"), ]
```

```{r}
train_xts <- xts(x = train$shipping_liters, order.by = train$`End of Week`)
test_xts <- xts(x = test$shipping_liters, order.by = test$`End of Week`)
```

```{r}
plot(train_xts)
```

```{r}
plot(test_xts)
```

Testing the first hypothesis -

```{r}
# Generating lagged variables to explore time-delayed relationships
for (i in 1:30) {
  train_lag <- train %>%
    mutate(lagged_variable = lag(train$shipping_liters, i))
  print(paste("Lag", i, "correlation:", cor(train_lag$consumption_liters, train_lag$lagged_variable, use = "complete.obs")))
}
```

The correlation coefficients for 30 lags range between 0.086 to 0.092, indicating weak linear relationships between past values of shipping and current values of consumption.

From here on we are focusing on the shipping dataset.

```{r}
shipping_train <- vodka_shipments[vodka_shipments$`End of Week` <= as.Date("2023-12-31"), ]
shipping_test <- vodka_shipments[vodka_shipments$`End of Week` >= as.Date("2024-01-01"), ]
```

```{r}
shipping_train_xts <- xts(x = shipping_train$shipping_liters, order.by = shipping_train$`End of Week`)
shipping_test_xts <- xts(x = shipping_test$shipping_liters, order.by = shipping_test$`End of Week`)
```

#### Holt Winters model for weekly data

```{r}
shipping_train_timeseries <- ts(shipping_train$shipping_liters, frequency = 52)
```

```{r}
holtwinters_shipping <- HoltWinters(shipping_train_timeseries)
holtwinters_shipping
```

```{r}
predictions_holtwinters_shipping <- forecast(holtwinters_shipping, h = nrow(shipping_test_xts))
predictions_holtwinters_shipping$mean
```

```{r}
accuracy(predictions_holtwinters_shipping$mean, shipping_test_xts)
```

```{r}
# Extract the time period for the forecast
forecast_dates <- seq(start(shipping_test_xts)[1], length = length(predictions_holtwinters_shipping$mean), by = "week")

# data frame to plot
hw_forecast_df <- data.frame(
  Date = forecast_dates,
  Forecast = predictions_holtwinters_shipping$mean,
  Actual = coredata(shipping_test_xts)
)

# Plotting
ggplot(data = hw_forecast_df, aes(x = Date)) +
  geom_line(aes(y = Actual, colour = "Actual"), size = 1.2) +
  geom_line(aes(y = Forecast, colour = "Forecast"), linetype = "dashed", size = 1.2) +
  labs(
    title = "Holt-Winters Weekly Forecast vs Actual",
    x = "Date",
    y = "Shipping Volume"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "red")) +
  theme_minimal()
```

#### Holt Winters model for adjusted seasonal data

```{r}
shipping_train_timeseries_2 <- ts(shipping_train$shipping_liters, frequency = 9 * 12)
```

```{r}
holtwinters_shipping_2 <- HoltWinters(shipping_train_timeseries_2)
holtwinters_shipping_2
```

```{r}
predictions_holtwinters_shipping_2 <- forecast(holtwinters_shipping_2, h = nrow(shipping_test_xts))
predictions_holtwinters_shipping_2$mean
```

```{r}
accuracy(predictions_holtwinters_shipping_2$mean, shipping_test_xts)
```

```{r}
adjusted_forecast_dates <- seq(start(shipping_test_xts)[1], length = length(predictions_holtwinters_shipping_2$mean), by = "week")

# data frame to plot
adjusted_hw_forecast_df <- data.frame(
  Date = adjusted_forecast_dates,
  Forecast = predictions_holtwinters_shipping_2$mean,
  Actual = coredata(shipping_test_xts)
)

# Plotting
ggplot(data = adjusted_hw_forecast_df, aes(x = Date)) +
  geom_line(aes(y = Actual, colour = "Actual"), size = 1.2) +
  geom_line(aes(y = Forecast, colour = "Forecast"), size = 1.2) +
  labs(
    title = "Adjusted Holt-Winters Forecast vs Actual",
    x = "Date",
    y = "Shipping Volume"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "red")) +
  theme_minimal()
```

```{r}
# Data frame for residuals
adjusted_hw_forecast_df <- data.frame(
  Date = index(shipping_test_xts),
  Forecast = predictions_holtwinters_shipping_2$mean,
  Actual = coredata(shipping_test_xts)
)

# Calculate residuals
adjusted_hw_forecast_df$Residuals <- adjusted_hw_forecast_df$Actual - adjusted_hw_forecast_df$Forecast

# Plotting residuals
library(ggplot2)
ggplot(data = adjusted_hw_forecast_df, aes(x = Date, y = Residuals)) +
  geom_line(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals of Adjusted Holt-Winters Forecast",
    x = "Date",
    y = "Residuals"
  ) +
  theme_minimal()
```

#### Multi-seasonal Time Series Analysis

```{r}
multiseasonal_time_series <- msts(shipping_train_timeseries_2, seasonal.periods = c(12, 9, 469))
plot(multiseasonal_time_series)
```

```{r}
tbats_model <- tbats(multiseasonal_time_series)
tbats_model
```

```{r}
tbats_components <- tbats.components(tbats_model)
plot(tbats_components)
```

```{r}
plot(forecast(tbats_model, h = 19))
```

```{r}
predictions_tbats_shipping <- forecast(tbats_model, h = nrow(shipping_test_xts))
predictions_tbats_shipping$mean
```

```{r}
accuracy(predictions_tbats_shipping$mean, shipping_test_xts)
```

```{r}
# Data frame to plot
tbats_forecast_df <- data.frame(
  Date = index(shipping_test_xts),
  Forecast = predictions_tbats_shipping$mean,
  Actual = coredata(shipping_test_xts)
)

names(tbats_forecast_df)[names(tbats_forecast_df) == "x"] <- "Forecast"

# Plotting
ggplot(data = tbats_forecast_df, aes(x = Date)) +
  geom_line(aes(y = Actual, colour = "Actual"), size = 1.2) +
  geom_line(aes(y = Forecast, colour = "Forecast"), size = 1.2) +
  labs(
    title = "TBATS Forecast vs Actual",
    x = "Date",
    y = "Shipping Volume"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "red")) +
  theme_minimal()
```

#### Combined Holt Winters + TBATS

```{r}
holtwinters_time_series <- ts(predictions_holtwinters_shipping_2$mean, start = start(shipping_test_xts), frequency = frequency(shipping_test_xts))
tbats_time_series <- ts(predictions_tbats_shipping$mean, start = start(shipping_test_xts), frequency = frequency(shipping_test_xts))

# Merge forecasts
combined_forecast <- merge(holtwinters_time_series, tbats_time_series, all = TRUE)
```

```{r}
# Data frame to plot
combined_forecast_df <- data.frame(
  Date = index(shipping_test_xts),
  Forecast = coredata(combined_forecast),
  Actual = coredata(shipping_test_xts)
)

names(combined_forecast_df)[names(combined_forecast_df) == "x"] <- "Forecast"

# Plotting
ggplot(data = combined_forecast_df, aes(x = Date)) +
  geom_line(aes(y = Actual), colour = "blue", size = 1.2) +
  geom_line(aes(y = Forecast), colour = "red", size = 1.2) +
  labs(
    title = "Combined Holt-Winters and TBATS Forecast vs Actual",
    x = "Date",
    y = "Shipping Volume"
  ) +
  theme_minimal()
```

```{r}
# Calculating residuals
combined_forecast_df <- data.frame(
  Date = index(shipping_test_xts),
  Forecast = coredata(combined_forecast),
  Actual = coredata(shipping_test_xts)
)

names(combined_forecast_df)[names(combined_forecast_df) == "x"] <- "Forecast"

combined_forecast_df$Residuals <- combined_forecast_df$Actual - combined_forecast_df$Forecast

# Plotting residuals
ggplot(data = combined_forecast_df, aes(x = Date, y = Residuals)) +
  geom_line(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals of Combined Holt-Winters and TBATS Forecast",
    x = "Date",
    y = "Residuals"
  ) +
  theme_minimal()
```

#### Combined Holt Winters + SARIMA

```{r}
shipping_train_timeseries_3 <- ts(shipping_train$shipping_liters, frequency = 12)
sarima_shipping <- auto.arima(shipping_train_timeseries_3, seasonal = TRUE)
summary(sarima_shipping)
```

```{r}
predictions_sarima_shipping <- forecast(sarima_shipping, h = nrow(shipping_test_xts))
predictions_sarima_shipping$mean
```

```{r}
holtwinters_time_series <- ts(predictions_holtwinters_shipping_2$mean, start = start(shipping_test_xts), frequency = frequency(shipping_test_xts))
sarima_timeseries <- ts(predictions_sarima_shipping$mean, start = start(shipping_test_xts), frequency = frequency(shipping_test_xts))

# Merge forecasts
combined_forecast <- merge(holtwinters_time_series, sarima_timeseries, all = TRUE) # Use common time points only

# Average the forecasts
combined_hw_sarima <- rowMeans(combined_forecast)
accuracy(combined_hw_sarima, shipping_test_xts)
```

```{r}
# data frame to plot
combined_sarima_df <- data.frame(
  Date = index(shipping_test_xts),
  Forecast_HW_SARIMA = combined_hw_sarima,
  Actual = coredata(shipping_test_xts)
)

# Plotting
ggplot(data = combined_sarima_df, aes(x = Date)) +
  geom_line(aes(y = Actual, colour = "Actual"), size = 1.2) +
  geom_line(aes(y = Forecast_HW_SARIMA, colour = "Forecast HW SARIMA"), size = 1.2) +
  labs(title = "Combined Holt-Winters and SARIMA Forecast vs Actual",
       x = "Date",
       y = "Shipping Volume") +
  scale_color_manual(values = c("Actual" = "blue", "Forecast HW SARIMA" = "red")) +
  theme_minimal()

```

```{r}
# Data frame for residuals
combined_hw_sarima_df <- data.frame(
  Date = index(shipping_test_xts),
  Forecast = combined_hw_sarima,
  Actual = coredata(shipping_test_xts)
)

# Calculate residuals
combined_hw_sarima_df$Residuals <- combined_hw_sarima_df$Actual - combined_hw_sarima_df$Forecast

# Plotting residuals
library(ggplot2)
ggplot(data = combined_hw_sarima_df, aes(x = Date, y = Residuals)) +
  geom_line(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals of Combined Holt-Winters and SARIMA Forecast",
    x = "Date",
    y = "Residuals"
  ) +
  theme_minimal()
```
